---
title: "Task 1. Exploring Stream Flow"
author: "John Fay"
date: " Fall 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
 1. Prepping your workspace for analysis
 2. Reading the data into an R dataframe
 3. Dataframes in R
 4. Exploratory data analysis: Plotting
 5. Exploratory data analysis: Summary statistics
 6. Comparing streamflow before and after Falls Lake dam constuction
 
# 1. Prepping your workspace for analysis

```{r Installing and loading libraries}
#Install the tidyverse package onto your machine
#install.packages('tidyverse')

#Load the tidyverse package into this session
library(tidyverse)

```

# 2. Reading the data into an R dataframe
 * See https://readr.tidyverse.org/reference/read_delim.html for more information
```{r Get the data}
#Get the URL 
url = 'http://waterservices.usgs.gov/nwis/dv/?format=rdb&sites=02087500&startDT=1930-10-01&endDT=2017-09-30&statCd=00003&parameterCd=00060&siteStatus=all'
#Read the data (tab separated) into a dataframe, skipping metadata lines, and assigning column names
theData <- read_tsv(file = url,
                   skip=30,
                   col_names = c('agency_cd','site_no','datetime','MeanFlow_cfs','Confidence'),
                   )
```

# 3. Dataframes in R
## 3a. Examples of dataframe properties and operations.
```{r Explore dataframes}
#Display the data (first 5 rows)
head(theData)

#Display the entire dataframe in a separate viewer
View(theData)

#Display the column names
colnames(theData)

#Display the structure of the dataframe
str(theData)

#Display the dimensions of the dataframe
dim(theData)

#Display the first 10 records of a single column of data
theData$MeanFlow_cfs[1:10]

#Display a summary of the data
summary(theData)

#Display summary of one data column
summary(theData$MeanFlow_cfs)

```



## 3c. Deriving new columns
 * Add a column for water flow in CMS them MGD
```{r Calculate CMS from CFS}
#Create cms column to plot cubic meters per second (1 cfs = 0.028316847 cms)
theData$MeanFlow_cms <- theData$MeanFlow_cfs*0.028316847
  
#Create MGD column to plot million gallons per day (1 cfs = 0.6464 mgd)
theData$MeanFlow_MGD <- theData$MeanFlow_cfs*0.6464

head(theData)
```

 * Extract month and year columns with `lubridate` (https://lubridate.tidyverse.org/)
```{r}
#Import the lubridate package
library(lubridate)

#Extract the month from the datetime column
theData$month <- month(theData$datetime)

#Extract the year from the datetime column
theData$year <- year(theData$datetime)

#What data types to month and year take?
str(theData)
```
 

## 3c.Using masks to update select rows: water year
Recall that water years start in October and end in September. So any rows occurring in Oct, Nov, or Dec should have a water year of year + 1. We do this using a mask
```{r Selecting rows of data}
theData$water.year <- theData$year 
theData$water.year[theData$month > 9] <- theData$year + 1
View(theData)
```

# 4. Exploratory data analysis - Plotting: 

## 4a. Basic plotting in R

* Basic Plots in R:
 * Create the margins of your plot using par
 * par(mar) = c(3,5,2,5)) where the margin is set for (bottom, left, top, right)
 * Plot the data… there are a lot of additions you can add to the plot with a ,
 * Basic plot: plot(xaxis, yaxis), so in this case: plot(neuse$Date, neuse$Flow_cms)
 * type tells the plot what type of plot you want. n tells the plot not to show points
        yaxt and xaxt tell the plot what to do with the y and x axis
        ylim and xlim tell the plot the min and max of the respective axes
        ylab and xlab set the labels for the respective axes. main sets the title for the plot.
    Currently, the plot is empty. The following commands fill the plot.
        axis lets you set up with more control the axis of interest. 1 is the x-axis. 2 is the y-axis.
        points() and lines() allow you to overlay data onto the plot as points or lines. Because there are so many points we want to plot the data as lines. There are many ways to stylize points and lines.
        col sets the color, lwd sets the width, lty sets the type of line
        abline allows you to draw straight lines in the plot. v for vertical lines, h for horizontal lines. Here we draw a vertical line around the time falls lake was constructed

```{r Basic plotting}
#Create margins for the plot
par(mar) = c(3,5,2,5) 

#Construct the plot template 
plot(theData$datetime,     #x-axis data
     theData$MeanFlow_cms, #y-axis data
     type='n',             #n = don't show points
     yaxt="n", 
     xlim=c(min(theData$datetime),
            max(theData$datetime)),
     ylab="Streamflow (cms)", 
     xlab = '', 
     main="Neuse River near Clayton, NC")

#Fill the template with data
axis(2, las=2, cex.axis=0.9) #Set the y-axis (#2)

#Add the mean flow line
lines(theData$datetime, 
      theData$MeanFlow_cms, 
      col="blue", 
      lty=1, 
      lwd=1)

#Add a vertical line where the dam was completed. 
abline(v=as.Date("1984-01-01"), 
       lty=3, 
       col="black", 
       lwd=2)
```

## 4b. Plotting using ggplot
```{r Plot using ggplot2}
#Import the library
library(ggplot)

#Construct the plot
ts <- ggplot(data=theData, aes(datetime,MeanFlow_cms)) +
      geom_line(colour="blue")

#make it prettier
ts <- ts + xlab("") + ylab("Streamflow (cms)") +
      ggtitle("Neuse River near Clayton, NC")

#show the plot
ts
```

# 5. Exloratory Data Analysis: Summary Statistics
#How confident are we in the data?
```{r Data confidence code}
#
total.n <- dim(theData)[1]; total.n
```